Parallel Adaptive Importance Sampling

### 2016-06-20 PAIS.tex
Final? edits.

However, we argue that this is not an optimal
scenario. First of all, unless we have a lot of information about the
posterior, we will initialise the algorithm's initial state in the
tails of the distribution. The samples that are initially made as the
algorithm finds its way to the mode(s) of the distribution cannot be
considered to be samples from the target distribution, and must be thrown
away. This process is known as the burn-in. In a trivially parallelised
scenario, each ensemble member must perform this process independently,
and therefore mass parallelisation makes no inroads to cutting this cost.

@@

Moreover, a trivially parallelised MCMC scheme is exactly that -
trivial. Intuition suggests that we can gain some speed up by sharing
information across the ensemble members, and this is
what we wish to demonstrate in this paper.

These deficiencies of the trivial method of parallelising MCMC methods
motivated the development of the Parallel Adaptive Importance Sampler
(PAIS). In the next section we will introduce the method in its most
general form.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.65\textwidth]{"figures/C1_burnin"}
\caption{The effective sample size ratio and relative $L^2$ error between the proposal and posterior distributions at each of the first 30 iterations. These numerics are taken from a simulation of the problem in Section~\ref{sec:chem} using the PAIS-Gamma algorithm.}
\label{fig:neff-burnin}
\end{figure}

Figure~\ref{fig:neff-burnin} shows that the effective sample size reaches stationarity at the same time as the relative $L^2$ error between the posterior distribution and the proposal distribution reaches its minimum. Detecting this stationarity allows us to automatically determine the end of the burn-in phase of the algorithm.

@@

There are of course infinitely many different ways that one could use
to split the original multinomial up into $M$ parts, some of which
will be far from optimal. The method that we have arrived at is based
on the idea that we want to group together states which are
geographically close to each other, preferentially doing this 
for the states with the biggest weights.

\begin{table}
\begin{mdframed}
\begin{algorithmic}
\STATE $\b{z} = M\bar{\b{w}}$
\FOR{$i=1,2,\ldots,M$}
\STATE $J = \argmax_j z_j$
\STATE $p_{i,J} = \min\{1,z_J\}$
\STATE $z_J = z_J - p_{i,J}$
\WHILE{$\sum_j p_{i,j} <1$}
\STATE $K = \argmin_{k \in \{k|z_k>0\}} \|y_J - y_k\|$
\STATE $p_{i,K} = \min\{1-\sum_j p_{i,j}, z_K\}$
\STATE $z_K = z_K - p_{i,K}$
\ENDWHILE
\STATE $x_i = \sum_k p_{i,k}y_k$.
\ENDFOR
\end{algorithmic}
\end{mdframed}
\caption{The approximate multinomial resampler (AMR) algorithm.\label{tab:AMR}}
\end{table}

Table \ref{tab:AMR} describes the basis of the algorithm with
deterministic resampling, using the means of each of the
sub-multinomials as the new samples. This resampler was designed with the aims of being
more efficient than the ETPF, and more accurate than straight multinomial
resampling. Therefore we now present numerical examples which
demonstrate this.

@@

We have explored the application of parallelised MCMC algorithms in
low dimensional inverse problems. We have demonstrated numerically
that these algorithms converge faster than the analogous naively parallelised
Metropolis-Hastings algorithms. Further experimentation with the
Random Walk Metropolis-Hastings (RWMH), Metropolis Adjusted Langevin
Algorithm (MALA), preconditioned Crank-Nicolson (pCN) and Hamiltonian
Monte Carlo (HMC) proposals has
yielded similar results\cite{Paul}.

>>>>

However, we argue that this is not an optimal
scenario. First of all, unless we have a lot of information about the
posterior, we will initialise the algorithm's initial state in the
tails of the distribution. The samples that are initially made as the
algorithm finds its way to the mode(s) of the distribution cannot be
considered to be samples from the target distribution, and must be thrown
away. This process is known as the burn-in. In a naively parallelised
scenario, each ensemble member must perform this process independently,
and therefore mass parallelisation makes no inroads to cutting this cost.

@@

Moreover, a naively parallelised MCMC scheme is exactly that -
naive. Intuition suggests that we can gain some speed up by sharing
information across the ensemble members, and this is
what we wish to demonstrate in this paper.

These deficiencies of the naive method of parallelising MCMC methods
motivated the development of the Parallel Adaptive Importance Sampler
(PAIS). In the next section we will introduce the method in its most
general form.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.65\textwidth]{"figures/C1_burnin"}
\caption{The effective sample size ratio and relative $L^2$ difference
between the proposal and posterior distributions at each of the first 30
iterations. These numerics are taken from a simulation of the problem in
Section~\ref{sec:chem} using the PAIS-Gamma algorithm.}
\label{fig:neff-burnin}
\end{figure}

Figure~\ref{fig:neff-burnin} shows that the effective sample size flattens
out at the same time as the relative $L^2$ difference between
the posterior distribution and the proposal distribution stabilises close to
its minimum. Detecting this stationarity allows us to automatically
determine the end of the burn-in phase of the algorithm.

@@

There are of course infinitely many different ways that one could use
to split the original multinomial up into $M$ parts, some of which
will be far from optimal. The method that we have chosen is loosely
based on the idea of optimal transport. We search out states with the
largest weights, and choose a cluster around these points based on
the closest states geographically. This method is not optimal since
once most of the clusters have been selected the remaining states
may be spread across the parameter space.

\begin{table}[htb]
\begin{mdframed}
\begin{algorithmic}
\STATE $\b{z} = M\bar{\b{w}}$
\FOR{$i=1,2,\ldots,M$}
\STATE $J = \argmax_j z_j$
\STATE $p_{i,J} = \min\{1,z_J\}$
\STATE $z_J = z_J - p_{i,J}$
\WHILE{$\sum_j p_{i,j} <1$}
\STATE $K = \argmin_{k \in \{k|z_k>0\}} \|y_J - y_k\|$
\STATE $p_{i,K} = \min\{1-\sum_j p_{i,j}, z_K\}$
\STATE $z_K = z_K - p_{i,K}$
\ENDWHILE
\STATE $x_i = \sum_k p_{i,k}y_k$.
\ENDFOR
\end{algorithmic}
\end{mdframed}
\caption{The approximate multinomial resampler (AMR) algorithm.\label{tab:AMR}}
\end{table}

Table \ref{tab:AMR} describes the basis of the algorithm with
deterministic resampling, using the means of each of the
sub-multinomials as the new samples. This resampler was designed with the aims of being
numerically cheaper than the ETPF, and more accurate than straight multinomial
resampling. Therefore we now present numerical examples which
demonstrate this.

@@

We have explored the application of parallelised MCMC algorithms in
low dimensional inverse problems. We have demonstrated numerically
that these algorithms converge faster than the analogous naively parallelised
Metropolis-Hastings algorithms. Further experimentation with the Metropolis
Adjusted Langevin Algorithm (MALA), preconditioned Crank-Nicolson (pCN),
preconditioned Crank-Nicolson Langevin (pCNL) and Hamiltonian
Monte Carlo (HMC) proposals has yielded similar results\cite{Paul}.
